---
title: "Forecasting future movement from speed trajectories in Roe Deer using a linear predictive model"
subtitle: "Data challenge :GEO880, Patterns and Trends in Environmental Data - Computational Movement Analysis"
author: "Tim Fässler and Gregory Biland"
date: "30/05/2022"
output:
  html_document:
    theme: spacelab
    highlight: monochrome
    toc: true
    toc_float: true
    number_sections: true
    text-align: justify
---

```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
``` 


```{r, message=FALSE, warning=FALSE,, include=FALSE}
Sys.setenv(LANG = "en")
library(readr)  # to import tabular data (e.g. csv)
library(tidyr)
library(tibble)
library(dplyr)        # to manipulate (tabular) data
library(sf)           # to handle spatial vector data
library(lubridate)    # To handle dates and times
library(ggplot2)
library(caTools)
library(caret)
library(raster)
library(prophet)
library(MASS)
library(jtools)
library(leaflet)
library(simplevis)

library(tseries)
library(ggfortify)
library(changepoint)
library(strucchange)
library(feasts)
library(forecast)
library(CINNA)
library(xts)
library(class)
library(caret)
library(moveHMM)


knitr::opts_chunk$set(
  message = F,
  fig.width = 7,
  fig.height = 6,
  pandoc.stack.size = "4g",
  fig.align = 'center',
  opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
)

dataFolder   <- here::here()   
RFolder      <- here::here()         
figureFolder <- here::here("figs") 
```

```{r, include=FALSE}

deer <- read_delim(file.path(dataFolder, "all_deer_complete.csv")) %>% data.frame()

canton_zh <- st_read(file.path(dataFolder,"Gemeindegrenzen_-OGD/UP_KANTON_F.shp"))
ASTK_ZH <- read_delim(file.path(dataFolder, "ASTK_ZH.csv"))
# ASTK <- read_delim(file.path(dataFolder, "Arealstatistik.csv"))
```

<div class=text-justify>

# Abstract  {.tabset}
Animal movement has always been a complex field of study as it consists of and compiles the difficulties of ecology, geography and data science. This project seeks to merge these three disciplines as it tries to predict the future movement of Roe Deer through space with the usage of an linear model with constraints set up by typical biological behaviours of the Roe Deer.

# Background and research goals  {.tabset}

With the continuous compaction of the soil and the ongoing sealing of the ground, the habitat of native species in Switzerland is increasingly under pressure [reference]. To counteract this pressure, it has become essential to sustainably protect animal habitats and thus preserve the diverse biodiversity. For this to be possible, it is important to know where the animals move, but also where possible areas are in which they may reside. This prediction of the habitats is the content of this work based on the GPS data of 15 tracked red deer in Zurich. Using GPS data from 15 deer in the south of the canton of Zurich, we evaluate their individual movement paths of three animals and, using a time model, try to predict the speed at which the animals will travel 24 hours from the last measurement point and thus determine the maximum distance the animals will travel. In order to be able to control how much the ground sourcing affects this prediction, a linear model was created to investigate the relationship between the area statistics and the speed.

Our research questions are:

- RQ.1: Using velocity, how accurately can movements in the future be modeled using a predictive time model?

- RQ.2: What parameters are crucial for modeling and predicting deer movement ranges using a linear predictive model (LM), and how much does the "Arealstatistik" influence this model?

# Methods and data  {.tabset}

## Data

```{r}
head(deer)
```

## Preprocessing
```{r}
#deer <- deer %>% group_by(reh) %>% mutate(Name = randomNames(reh, ethnicity = 4)) #Give deer names
canton_zh <- canton_zh %>% subset(KANTON == "Zürich")
canton_zh <- canton_zh %>% st_transform(2056)

# ASTK <- ASTK %>% st_as_sf(coords = c("E", "N"), crs = 2056, remove = FALSE)
# ASTK_ZH <- st_join(canton_zh, ASTK, join = st_contains_properly, left=TRUE)
# ASTK_ZH <- ASTK_ZH %>% st_drop_geometry()
# write.csv(ASTK_ZH, file = "ASTK_ZH.csv", row.names=FALSE)

# Sex to binary: male = 0, female = 1, rename "reh" to "TierID"
deer$sex[deer$sex == "m"] <- as.numeric(0)
deer$sex[deer$sex == "f"] <- as.numeric(1)

deer <- deer %>% rename(TierID = reh, DatetimeUTC = datetime_utc)
colnames(deer)[c(3,4)] <- c("x","y")

deer <- deer %>% mutate(
  x = x + 2000000,
  y = y + 1000000
)

# Making a bounding box of the relevant study area
deer <- deer %>% st_as_sf(coords = c("x", "y"), crs = 2056, remove = FALSE) 
bbox <- st_make_grid(st_bbox(deer), n = 1)
bbox_df <- bbox[[1]][[1]] %>% data.frame()
deer <- deer %>% st_drop_geometry()

# Making the bounding box a bit bigger, so that our predictions can go further outside, 1km on every side
bbox_df$X1[1] = bbox_df$X1[1] - 1000
bbox_df$X1[2] = bbox_df$X1[2] + 1000
bbox_df$X1[3] = bbox_df$X1[3] + 1000
bbox_df$X1[4] = bbox_df$X1[4] - 1000
bbox_df$X1[5] = bbox_df$X1[5] - 1000

bbox_df$X2[1] = bbox_df$X2[1] - 1000
bbox_df$X2[2] = bbox_df$X2[2] - 1000
bbox_df$X2[3] = bbox_df$X2[3] + 1000
bbox_df$X2[4] = bbox_df$X2[4] + 1000
bbox_df$X2[5] = bbox_df$X2[5] - 1000
bbox_sf <- bbox_df %>% st_as_sf(coords = c('X1', 'X2'), crs = 2056)

bbox_sf <- bbox_sf %>% 
  summarise(geometry = st_combine(geometry)) %>%
  st_cast("POLYGON")

bbox_sf <- bbox_sf %>% st_transform(2056)

```

```{r}
#Rename deer as numeric and remove long lat
deer <- deer %>% mutate(sex = as.numeric(sex))

```

```{r}
#Plot data for overview
deer_sf_LL <- deer %>% st_as_sf(coords = c("longitude", "latitude"), crs = 2056, remove = FALSE)
deer_sf <- deer %>% st_as_sf(coords = c("x", "y"), crs = 2056, remove = FALSE)
ggplot()+
  geom_sf(data=bbox_sf, alpha=.1)+
  geom_path(data = deer,aes(x,y,  col = TierID))+
  geom_point(data = deer,aes(x,y, col = TierID))+
  ggtitle("Movement patterns for 15 different roe deer with the bounding box for \nthe observation field around")+
  xlab("X")+
  ylab("Y")+
  theme_minimal() +
  coord_sf(datum = 2056)

# leaf_sf_col(deer_sf_LL, 
#                col_var = TierID)
```

## Geometry calculations 

```{r}
deer <- deer %>% group_by(TierID) %>% mutate(
  timelag = as.integer(difftime(lead(DatetimeUTC),DatetimeUTC, units = 'secs')), #timelag is in seconds
  steplength = sqrt((y - lead(y,1))^2+ (x - lead(x,1))^2))

deer <- deer %>% mutate(speed = (steplength / timelag)) #speed is in meter per minute
deer <- deer %>% drop_na() %>% data.frame() %>% subset(select= -c(longitude, latitude) )
```

## Outlier Handling

```{r}

deer_quantiles_speed <- deer %>% group_by(TierID) %>% summarize(qua1 = quantile(speed, 0.01, na.rm = TRUE), qua2 = quantile(speed, 0.99, na.rm = TRUE))

deer <- deer %>% split(deer$TierID)

deer$RE01 <- deer$RE01[deer$RE01$speed >= deer_quantiles_speed$qua1[1] & 
                       deer$RE01$speed <= deer_quantiles_speed$qua2[1],]
deer$RE02 <- deer$RE02[deer$RE02$speed >= deer_quantiles_speed$qua1[2] & 
                       deer$RE02$speed <= deer_quantiles_speed$qua2[2],]
deer$RE03 <- deer$RE03[deer$RE03$speed >= deer_quantiles_speed$qua1[3] & 
                       deer$RE03$speed <= deer_quantiles_speed$qua2[3],]
deer$RE04 <- deer$RE04[deer$RE04$speed >= deer_quantiles_speed$qua1[4] & 
                       deer$RE04$speed <= deer_quantiles_speed$qua2[4],]
deer$RE05 <- deer$RE05[deer$RE05$speed >= deer_quantiles_speed$qua1[5] & 
                       deer$RE05$speed <= deer_quantiles_speed$qua2[5],]
deer$RE06 <- deer$RE06[deer$RE06$speed >= deer_quantiles_speed$qua1[6] & 
                       deer$RE06$speed <= deer_quantiles_speed$qua2[6],]
deer$RE07 <- deer$RE07[deer$RE07$speed >= deer_quantiles_speed$qua1[7] & 
                       deer$RE07$speed <= deer_quantiles_speed$qua2[7],]
deer$RE08 <- deer$RE08[deer$RE08$speed >= deer_quantiles_speed$qua1[8] & 
                       deer$RE08$speed <= deer_quantiles_speed$qua2[8],]
deer$RE09 <- deer$RE09[deer$RE09$speed >= deer_quantiles_speed$qua1[9] & 
                       deer$RE09$speed <= deer_quantiles_speed$qua2[9],]
deer$RE010 <- deer$RE010[deer$RE010$speed >= deer_quantiles_speed$qua1[10] & 
                       deer$RE010$speed <= deer_quantiles_speed$qua2[10],]
deer$RE011 <- deer$RE011[deer$RE011$speed >= deer_quantiles_speed$qua1[11] & 
                       deer$RE011$speed <= deer_quantiles_speed$qua2[11],]
deer$RE012 <- deer$RE012[deer$RE012$speed >= deer_quantiles_speed$qua1[12] & 
                       deer$RE012$speed <= deer_quantiles_speed$qua2[12],]
deer$RE013 <- deer$RE013[deer$RE013$speed >= deer_quantiles_speed$qua1[13] & 
                       deer$RE013$speed <= deer_quantiles_speed$qua2[13],]
deer$RE014 <- deer$RE014[deer$RE014$speed >= deer_quantiles_speed$qua1[14] & 
                       deer$RE014$speed <= deer_quantiles_speed$qua2[14],]
deer$RE015 <- deer$RE015[deer$RE015$speed >= deer_quantiles_speed$qua1[15] & 
                       deer$RE015$speed <= deer_quantiles_speed$qua2[15],]

deer_cleaned <- do.call(rbind, deer)

# Recalculating timelag, steplength and speed

deer_cleaned <- deer_cleaned %>% group_by(TierID) %>% mutate(
  timelag = as.integer(difftime(lead(DatetimeUTC),DatetimeUTC, units = 'secs')), #timelag is in minutes
  steplength = sqrt((y - lead(y,1))^2+ (x - lead(x,1))^2)) %>% mutate(speed = (steplength / timelag))

# Calculating additional attribute
colnames(deer_cleaned)[3] <- "ID"
deer_cleaned <- prepData(deer_cleaned, type = "UTM", ) %>% 
  data.frame() %>% 
  subset(select = -step) %>% 
  relocate(angle, .after = speed) %>%
  rename(TierID = ID) %>%
  relocate(TierID, .before = sex)

```


```{r}
# So wärs mit 0.01 und 0.99 percentil removed, chönd au no meh filtere
ggplot()+
  geom_sf(data=bbox_sf, alpha=.1)+
  geom_path(data = deer_cleaned,aes(x,y,  col = TierID))+
  geom_point(data = deer_cleaned,aes(x,y, col = TierID))+
  ggtitle("Movement patterns for 15 different roe deer with the bounding box for the observation field around")+
  xlab("X")+
  ylab("Y")+
  theme_minimal() +
  coord_sf(datum = 2056)
```

```{r}
ggplot()+
  geom_line(data = deer_cleaned, aes(x=DatetimeUTC, y = speed, col= TierID))+
  ggtitle("Monitoring timespan for 15 different roe deer")+
  ylab("Speed [m/s]")+
  xlab("Deer")

ggplot()+
  geom_line(data = deer_cleaned, aes(x=DatetimeUTC, y = steplength, col= TierID))+
  ggtitle("Monitoring timespan for 15 different roe deer")+
  ylab("Steplegth [m]")+
  xlab("Deer")

ggplot()+
  geom_line(data = deer_cleaned, aes(x=DatetimeUTC, y = timelag, col= TierID))+
  ggtitle("Monitoring timespan for 15 different roe deer")+
  ylab("Steplegth [m]")+
  xlab("Deer")
```

## Arealstatistik

```{r}

# Updating bounding box
bbox_df$X2[c(1,2,5)] <- min(deer_cleaned$y) - 1000

bbox_sf <- bbox_df %>% st_as_sf(coords = c('X1', 'X2'), crs = 2056)
bbox_sf <- bbox_sf %>% 
  summarise(geometry = st_combine(geometry)) %>%
  st_cast("POLYGON")
bbox_sf <- bbox_sf %>% st_transform(2056)



# Extracting the relevant cells of the Arealstatistik
ASTK_ZH <- ASTK_ZH %>% st_as_sf(coords = c("E", "N"), crs = 2056, remove = FALSE)
ASTK_bbox <- st_join(bbox_sf, ASTK_ZH, join = st_contains_properly, left=TRUE)
ASTK_bbox <- ASTK_bbox %>% st_drop_geometry()

# Chosing which classification of the Arealstatistik we want to use
ASTK_bbox <- ASTK_bbox %>% dplyr::select(RELI, E, N, AS18_72)

# Making an empty raster with the extent of the Arealstatistik
empty_raster <- raster(
  ncols = (range(ASTK_bbox$E)[2]-range(ASTK_bbox$E)[1])/100+1, 
  nrows =  (range(ASTK_bbox$N)[2]-range(ASTK_bbox$N)[1])/100+1, 
  xmn = min(ASTK_bbox$E)-50, 
  xmx = max(ASTK_bbox$E)+50, 
  ymn = min(ASTK_bbox$N)-50,
  ymx = max(ASTK_bbox$N)+50
  )

# Populating the raster
ASTK_raster <- rasterize(dplyr::select(ASTK_bbox, E, N), empty_raster, ASTK_bbox$AS18_72) 

# Adding arealstatistik values to deer data
deer_cleaned <- data.frame(deer_cleaned)
deer_cleaned <- deer_cleaned %>% mutate(astk = as.vector(extract(ASTK_raster, st_as_sf(deer_cleaned, coords = c("x", "y")) , 'simple')))

deer_cleaned$astk[is.na(deer_cleaned$astk)] <- 0
deer_cleaned <- deer_cleaned %>% na.omit()
```

```{r}
raster_vis <- as.data.frame(ASTK_raster, xy = TRUE)

ggplot()+
  # geom_sf(data = canton_zh, fill = alpha("white", .1))+
  geom_sf(data = bbox_sf)+
  geom_raster(data = raster_vis, aes(x = x, y = y, alpha = layer))+
  geom_path(data = deer_cleaned, aes(x = x, y = y, col = TierID), alpha = 0.3)+
  geom_point(data = deer_cleaned, aes(x = x, y = y, col = TierID), size = .1)+
  coord_sf(datum = 2056)+
  theme_minimal()+
  ggtitle("Tracking points for roe deer")
  
ggplot()+
  geom_raster(data = raster_vis, aes(x = x, y = y, alpha = layer))+
  coord_sf(datum = 2056)+
  geom_point(data = deer_cleaned, aes(x,y, color = as.factor(astk)))+
  geom_path(data = deer_cleaned, aes(x = x, y = y), alpha = 0.3)+
  theme_minimal()+
  ggtitle("Tracking points for roe deer with the Arealstatistik")
```

## Methods

```{r}
# Filte roe deer 1,2 and 3 and change coordinates to sf object
deer_01 <- deer_cleaned %>% filter(TierID == "RE01") %>% na.omit()
deer_02 <- deer_cleaned %>% filter(TierID == "RE02") %>% na.omit()
deer_03 <- deer_cleaned %>% filter(TierID == "RE03") %>% na.omit()

deer1_3 <- rbind(deer_01, deer_02, deer_03)

deer_01_sf <- deer_01 %>% st_as_sf(coords = c("x", "y"), crs = 2056, remove = FALSE)
deer_02_sf <- deer_01 %>% st_as_sf(coords = c("x", "y"), crs = 2056, remove = FALSE)
deer_03_sf <- deer_01 %>% st_as_sf(coords = c("x", "y"), crs = 2056, remove = FALSE)

```

```{r}
#Overview over changepoints in speed for the roe deer
CP_value01 = cpt.mean(deer_01$speed, method= "BinSeg") # Using method "binseg" cause it is quick but only approximate
CP_value01
cpts(CP_value01)

plot(CP_value01,xlab="Date", ylab = "Speed [m/s]" ,main =" Change points in the speed of roe deer 1")

CP_value02 = cpt.mean(deer_02$speed, method= "BinSeg") # Using method "binseg" cause it is quick but only approximate
CP_value02
cpts(CP_value02)

plot(CP_value02,xlab="Date", ylab = "Speed [m/s]" ,main =" Change points in the speed of roe deer 2")

CP_value03 = cpt.mean(deer_03$speed, method= "BinSeg") # Using method "binseg" cause it is quick but only approximate
CP_value03
cpts(CP_value03)

plot(CP_value03,xlab="Date", ylab = "Speed [m/s]" ,main =" Change points in the speed of roe deer 3")
```

Surprisingly we do not set any changepoints in the speed values for the roe deer.

```{r}
### Forecasting speed 
speed_forcasting <- function(file){
  deer_mean_speed = file %>% 
  group_by(DatetimeUTC) %>% 
  summarise(speed = mean(speed, na.rm = TRUE)) %>% 
  ungroup() %>% 
  rename(ds = DatetimeUTC, y = speed)

m.s= prophet(df = deer_mean_speed)

future.s = make_future_dataframe(m = m.s, 
                               periods = 15, 
                               freq = "day")

future.speed = stats::predict(m.s, future.s)

print(plot(m.s,future.speed) +
  theme_bw()+
  theme(axis.text = element_text(size = 12), axis.title = element_text(size = 14))+
  labs(x = NULL, y = expression(Roe~deer~speed~(m/s))))

dyplot.prophet(x = m.s, fcst = future.speed,uncertainty = TRUE )

prophet_plot_components(m = m.s, fcst = future.speed)
}

deer_01_s.f <- speed_forcasting(deer_01)
deer_02_s.f <- speed_forcasting(deer_02)
deer_03_s.f <- speed_forcasting(deer_03)
```
## Home range estimation

```{r}
min_max <- function(file){ 
  min_speed = tail(file[[1]][["data"]][["trend_lower"]], n = 1) 
  max_speed = tail(file[[1]][["data"]][["trend_upper"]], n = 1)
  min_distance = (min_speed * 86400)
  max_distance = (max_speed * 86400)
  df <- data.frame (min_HR = min_speed,
                    max_HR = max_speed,
                    min_distance = min_distance,
                    max_distance = max_distance)
}

deer01_HR <- min_max(deer_01_s.f)
deer02_HR <- min_max(deer_02_s.f)
deer03_HR <- min_max(deer_03_s.f)

#Home range distance definition
homerange <- function(file, f1){
  last_point_x = last(file$x)
  last_point_y = last(file$y)
  
  buffer_point <- data.frame(x = last_point_x, y = last_point_y)
  buffer_point <- buffer_point %>% st_as_sf(coords = c("x", "y"), crs = 2056, remove = FALSE)
  
  min_buffer = buffer_point %>% st_buffer(dist= f1$min_distance)
  max_buffer = buffer_point %>% st_buffer(dist= f1$max_distance)
  rbind(min_buffer, max_buffer)
}

homerange_deer1 <- homerange(deer_01, deer01_HR)
homerange_deer2 <- homerange(deer_02, deer02_HR)
homerange_deer3 <- homerange(deer_03, deer03_HR)

ggplot()+
  geom_path(data = deer1_3, aes(x,y, col = TierID))+
  geom_point(data = deer1_3, aes(x,y, col = TierID))+
  geom_sf(data = homerange_deer1, col = "red", alpha = .1)+
  geom_sf(data = homerange_deer2, col = "darkgreen", alpha = .1)+
  geom_sf(data = homerange_deer3, col = "blue", alpha = .1)+
  ggtitle("Predicted homerange for Roe deer 1,2 and 3 in the next 24 hours \nwith the minimum and maximum distance")+
  xlab("X")+
  ylab("Y")+
  theme_minimal()+
  coord_sf(datum = 2056)
```

## Linear predictive model

```{r LM}
# https://cran.r-project.org/web/packages/jtools/vignettes/effect_plot.html
#Linear approach
fit <- lm(speed ~ sex + timelag + x + y + steplength + astk + angle, data = deer_cleaned)
fit_poly.astk <- lm(speed ~ sex + timelag + x + y + steplength + poly(astk, 2) + angle, data = deer_cleaned)
fit_poly.sex <- lm(speed ~ poly(sex, 1) + timelag + x + y + steplength + astk + angle, data = deer_cleaned)

#predicts the future values
s.astk.model <- lm(speed~astk,data = deer_cleaned)
s.sex.model <- lm(speed~sex,data = deer_cleaned)

variable_astk <-data.frame(astk=c(deer_cleaned$astk))
p1_astk <- predict(s.astk.model, newdata = variable_astk)
p2_astk <- predict(s.astk.model, newdata = variable_astk,interval = 'confidence')

variable_sex <-data.frame(sex=c(deer_cleaned$sex))
p1_sex <- predict(s.sex.model, newdata = variable_sex)
p2_sex <- predict(s.sex.model, newdata = variable_sex,interval = 'confidence')

# KNN- algorithm approach with prediction of speed
# Define training control
set.seed(54)

train.control <- trainControl(method = "cv", number = 10)

# Train the model
model <- train(speed ~ sex + timelag + x + y + steplength + astk+ angle
               , data = deer_cleaned
               , trControl = train.control, method = "knn")

# Summarize the results
predict_model_speed <- predict(model)
predict_model_speed <- predict_model_speed %>% data.frame()
predict_model_speed$row_num <- seq.int(nrow(predict_model_speed)) 
deer_cleaned$row_num <- seq.int(nrow(deer_cleaned)) 
```

# Results    {.tabset}

```{r}
#Summary
summ(fit)
summ(s.astk.model)
summ(s.sex.model)

autoplot(fit)
autoplot(fit_poly.astk)
autoplot(fit_poly.sex)

effect_plot(fit, pred = astk, interval = TRUE, plot.points = TRUE)
effect_plot(fit, pred = sex, interval = TRUE, plot.points = TRUE)
effect_plot(s.astk.model, pred = astk, interval = TRUE, plot.points = TRUE)
effect_plot(s.sex.model, pred = sex, interval = TRUE, plot.points = TRUE)

effect_plot(fit_poly.astk, pred = astk, interval = TRUE, plot.points = TRUE)
effect_plot(fit_poly.sex, pred = sex, interval = TRUE, plot.points = TRUE)

effect_plot(fit, pred = astk, interval = TRUE, plot.points = TRUE,
            jitter = .2)

```

The diagnostic plots show residuals in four different ways:

Residuals vs Fitted. Used to check the linear relationship assumptions. A horizontal line, without distinct patterns is an indication for a linear relationship, what is good.

Normal Q-Q. Used to examine whether the residuals are normally distributed. It’s good if residuals points follow the straight dashed line.

Scale-Location (or Spread-Location). Used to check the homogeneity of variance of the residuals (homoscedasticity). Horizontal line with equally spread points is a good indication of homoscedasticity. This is not the case in our example, where we have a heteroscedasticity problem.

Residuals vs Leverage. Used to identify influential cases, that is extreme values that might influence the regression results when included or excluded from the analysis. This plot will be described further in the next sections.


```{r}
plot(p1_astk)
plot(p2_astk)

plot(p1_sex)
plot(p2_sex)
```

```{r}
plot(model)
print(model)

ggplot()+
  geom_line(data = predict_model_speed, aes(x = row_num, y = .), col = "red", size = 1.5)+
  geom_line(data = deer_cleaned, aes(x = row_num, y = speed), col = "blue", size = .3)+
  xlab("Data point")+
  ylab("Speed [m/s]")+
  ggtitle("Predicted speed [red] using k-nearest neighboors algorithm plotted \nagainst observed speed [blue] in roe deer")

```

# Discussion    {.tabset}


<div>

# Bibliography
