---
title: "Forecasting coordinates from movement trajectories in Roe Deer"
subtitle: "Data challenge :GEO880, Patterns and Trends in Environmental Data - Computational Movement Analysis"
author: "Tim Fässler and Gregory Biland"
date: "30/05/2022"
output:
  html_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r, message=FALSE, warning=FALSE,, include=FALSE}
Sys.setenv(LANG = "en")
library(readr)  # to import tabular data (e.g. csv)
library(tidyr)
library(tibble)
library(dplyr)        # to manipulate (tabular) data
library(sf)           # to handle spatial vector data
library(lubridate)    # To handle dates and times
library(ggplot2)
library(caTools)
library(caret)
library(raster)
library(prophet)
library(MASS)
library(jtools)
library(leaflet)
library(simplevis)

library(tseries)
library(ggfortify)
library(changepoint)
library(strucchange)
library(feasts)
library(forecast)
library(CINNA)
library(xts)


knitr::opts_chunk$set(
  message = F,
  fig.width = 7,
  fig.height = 6,
  pandoc.stack.size = "4g",
  fig.align = 'center',
  opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
)

dataFolder   <- here::here()   
RFolder      <- here::here()         
figureFolder <- here::here("figs") 
```

```{r, include=FALSE}

deer <- read_delim(file.path(dataFolder, "all_deer_complete.csv")) %>% data.frame()

canton_zh <- st_read(file.path(dataFolder,"Gemeindegrenzen_-OGD/UP_KANTON_F.shp"))
ASTK_ZH <- read_delim(file.path(dataFolder, "ASTK_ZH.csv"))
# ASTK <- read_delim(file.path(dataFolder, "Arealstatistik.csv"))
```

## Abstract
Animal movement has always been a complex field of study as it consists of and compiles the difficulties of ecology, geography and data science. This project seeks to merge these three disciplines as it tries to predict the future movement of Roe Deer through space with the usage of an linear model with constraints set up by typical biological behaviours of the Roe Deer.

## Background and research goals
- What parameters are crucial for modeling and predicting deer movement patterns using an artificial neural network (ANN), and how much does the deers gender influence these models?

- Does the artificial neural network differentiate significantly between the deers movement in different kind of landscape spaces, for example forest vs on an open grass field.

## Methods and data

### Data
```{r}
#To test make subset of df
# deer <- subset(deer[1:100,])

head(deer)
```

### Preprocessing
```{r}
#deer <- deer %>% group_by(reh) %>% mutate(Name = randomNames(reh, ethnicity = 4)) #Give deer names
canton_zh <- canton_zh %>% subset(KANTON == "Zürich")
canton_zh <- canton_zh %>% st_transform(2056)

# ASTK <- ASTK %>% st_as_sf(coords = c("E", "N"), crs = 2056, remove = FALSE)
# ASTK_ZH <- st_join(canton_zh, ASTK, join = st_contains_properly, left=TRUE)
# ASTK_ZH <- ASTK_ZH %>% st_drop_geometry()
# write.csv(ASTK_ZH, file = "ASTK_ZH.csv", row.names=FALSE)

# Sex to binary: male = 0, female = 1, rename "reh" to "TierID"
deer$sex[deer$sex == "m"] <- as.numeric(0)
deer$sex[deer$sex == "f"] <- as.numeric(1)

deer <- deer %>% rename(TierID = reh, DatetimeUTC = datetime_utc)
colnames(deer)[c(3,4)] <- c("x","y")

deer <- deer %>% mutate(
  x = x + 2000000,
  y = y + 1000000
)

# Making a bounding box of the relevant study area
deer <- deer %>% st_as_sf(coords = c("x", "y"), crs = 2056, remove = FALSE) 
bbox <- st_make_grid(st_bbox(deer), n = 1)
bbox_df <- bbox[[1]][[1]] %>% data.frame()
deer <- deer %>% st_drop_geometry()

# Making the bounding box a bit bigger, so that our predictions can go further outside, 1km on every side
bbox_df$X1[1] = bbox_df$X1[1] - 1000
bbox_df$X1[2] = bbox_df$X1[2] + 1000
bbox_df$X1[3] = bbox_df$X1[3] + 1000
bbox_df$X1[4] = bbox_df$X1[4] - 1000
bbox_df$X1[5] = bbox_df$X1[5] - 1000

bbox_df$X2[1] = bbox_df$X2[1] - 1000
bbox_df$X2[2] = bbox_df$X2[2] - 1000
bbox_df$X2[3] = bbox_df$X2[3] + 1000
bbox_df$X2[4] = bbox_df$X2[4] + 1000
bbox_df$X2[5] = bbox_df$X2[5] - 1000
bbox_sf <- bbox_df %>% st_as_sf(coords = c('X1', 'X2'), crs = 2056)

bbox_sf <- bbox_sf %>% 
  summarise(geometry = st_combine(geometry)) %>%
  st_cast("POLYGON")

bbox_sf <- bbox_sf %>% st_transform(2056)

```

#Rename deer as numeric and remove long lat
```{r}
deer <- deer %>% mutate(sex = as.numeric(sex))

```


```{r}
#Plot data for overview
deer_sf_LL <- deer %>% st_as_sf(coords = c("longitude", "latitude"), crs = 2056, remove = FALSE)
deer_sf <- deer %>% st_as_sf(coords = c("x", "y"), crs = 2056, remove = FALSE)
ggplot()+
  geom_sf(data=bbox_sf, alpha=.1)+
  geom_path(data = deer,aes(x,y,  col = TierID))+
  geom_point(data = deer,aes(x,y, col = TierID))+
  ggtitle("Movement patterns for 15 different roe deer with the bounding box for the observation field around")+
  xlab("X")+
  ylab("Y")+
  theme_minimal() +
  coord_sf(datum = 2056)

# leaf_sf_col(deer_sf_LL, 
#                col_var = TierID)
```

### Geometry calculations
```{r}
deer <- deer %>% group_by(TierID) %>% mutate(
  timelag = as.integer(difftime(lead(DatetimeUTC),DatetimeUTC, units = 'secs')), #timelag is in seconds
  steplength = sqrt((y - lead(y,1))^2+ (x - lead(x,1))^2))

deer <- deer %>% mutate(speed = (steplength / timelag)) #speed is in meter per minute
deer <- deer %>% drop_na() %>% data.frame() %>% subset(select= -c(longitude, latitude) )
```

### Outlier Handling
```{r}

deer_quantiles_speed <- deer %>% group_by(TierID) %>% summarize(qua1 = quantile(speed, 0.01, na.rm = TRUE), qua2 = quantile(speed, 0.99, na.rm = TRUE))

deer <- deer %>% split(deer$TierID)

deer$RE01 <- deer$RE01[deer$RE01$speed >= deer_quantiles_speed$qua1[1] & 
                       deer$RE01$speed <= deer_quantiles_speed$qua2[1],]
deer$RE02 <- deer$RE02[deer$RE02$speed >= deer_quantiles_speed$qua1[2] & 
                       deer$RE02$speed <= deer_quantiles_speed$qua2[2],]
deer$RE03 <- deer$RE03[deer$RE03$speed >= deer_quantiles_speed$qua1[3] & 
                       deer$RE03$speed <= deer_quantiles_speed$qua2[3],]
deer$RE04 <- deer$RE04[deer$RE04$speed >= deer_quantiles_speed$qua1[4] & 
                       deer$RE04$speed <= deer_quantiles_speed$qua2[4],]
deer$RE05 <- deer$RE05[deer$RE05$speed >= deer_quantiles_speed$qua1[5] & 
                       deer$RE05$speed <= deer_quantiles_speed$qua2[5],]
deer$RE06 <- deer$RE06[deer$RE06$speed >= deer_quantiles_speed$qua1[6] & 
                       deer$RE06$speed <= deer_quantiles_speed$qua2[6],]
deer$RE07 <- deer$RE07[deer$RE07$speed >= deer_quantiles_speed$qua1[7] & 
                       deer$RE07$speed <= deer_quantiles_speed$qua2[7],]
deer$RE08 <- deer$RE08[deer$RE08$speed >= deer_quantiles_speed$qua1[8] & 
                       deer$RE08$speed <= deer_quantiles_speed$qua2[8],]
deer$RE09 <- deer$RE09[deer$RE09$speed >= deer_quantiles_speed$qua1[9] & 
                       deer$RE09$speed <= deer_quantiles_speed$qua2[9],]
deer$RE010 <- deer$RE010[deer$RE010$speed >= deer_quantiles_speed$qua1[10] & 
                       deer$RE010$speed <= deer_quantiles_speed$qua2[10],]
deer$RE011 <- deer$RE011[deer$RE011$speed >= deer_quantiles_speed$qua1[11] & 
                       deer$RE011$speed <= deer_quantiles_speed$qua2[11],]
deer$RE012 <- deer$RE012[deer$RE012$speed >= deer_quantiles_speed$qua1[12] & 
                       deer$RE012$speed <= deer_quantiles_speed$qua2[12],]
deer$RE013 <- deer$RE013[deer$RE013$speed >= deer_quantiles_speed$qua1[13] & 
                       deer$RE013$speed <= deer_quantiles_speed$qua2[13],]
deer$RE014 <- deer$RE014[deer$RE014$speed >= deer_quantiles_speed$qua1[14] & 
                       deer$RE014$speed <= deer_quantiles_speed$qua2[14],]
deer$RE015 <- deer$RE015[deer$RE015$speed >= deer_quantiles_speed$qua1[15] & 
                       deer$RE015$speed <= deer_quantiles_speed$qua2[15],]

deer_cleaned <- do.call(rbind, deer)

# Recalculating timelag, steplength and speed

deer_cleaned <- deer_cleaned %>% group_by(TierID) %>% mutate(
  timelag = as.integer(difftime(lead(DatetimeUTC),DatetimeUTC, units = 'secs')), #timelag is in minutes
  steplength = sqrt((y - lead(y,1))^2+ (x - lead(x,1))^2)) %>% mutate(speed = (steplength / timelag))
```


```{r}
# So wärs mit 0.01 und 0.99 percentil removed, chönd au no meh filtere
ggplot()+
  geom_sf(data=bbox_sf, alpha=.1)+
  geom_path(data = deer_cleaned,aes(x,y,  col = TierID))+
  geom_point(data = deer_cleaned,aes(x,y, col = TierID))+
  ggtitle("Movement patterns for 15 different roe deer with the bounding box for the observation field around")+
  xlab("X")+
  ylab("Y")+
  theme_minimal() +
  coord_sf(datum = 2056)
```



```{r}
ggplot()+
  geom_line(data = deer_cleaned, aes(x=DatetimeUTC, y = speed, col= TierID))+
  ggtitle("Monitoring timespan for 15 different roe deer")+
  ylab("Speed [m/s]")+
  xlab("Deer")

ggplot()+
  geom_line(data = deer_cleaned, aes(x=DatetimeUTC, y = steplength, col= TierID))+
  ggtitle("Monitoring timespan for 15 different roe deer")+
  ylab("Steplegth [m]")+
  xlab("Deer")
```

### Arealstatistik
```{r}

# Updating bounding box
bbox_df$X2[c(1,2,5)] <- min(deer_cleaned$y) - 1000

bbox_sf <- bbox_df %>% st_as_sf(coords = c('X1', 'X2'), crs = 2056)
bbox_sf <- bbox_sf %>% 
  summarise(geometry = st_combine(geometry)) %>%
  st_cast("POLYGON")
bbox_sf <- bbox_sf %>% st_transform(2056)



# Extracting the relevant cells of the Arealstatistik
ASTK_ZH <- ASTK_ZH %>% st_as_sf(coords = c("E", "N"), crs = 2056, remove = FALSE)
ASTK_bbox <- st_join(bbox_sf, ASTK_ZH, join = st_contains_properly, left=TRUE)
ASTK_bbox <- ASTK_bbox %>% st_drop_geometry()

# Chosing which classification of the Arealstatistik we want to use
ASTK_bbox <- ASTK_bbox %>% dplyr::select(RELI, E, N, AS18_72)

# Making an empty raster with the extent of the Arealstatistik
empty_raster <- raster(
  ncols = (range(ASTK_bbox$E)[2]-range(ASTK_bbox$E)[1])/100+1, 
  nrows =  (range(ASTK_bbox$N)[2]-range(ASTK_bbox$N)[1])/100+1, 
  xmn = min(ASTK_bbox$E)-50, 
  xmx = max(ASTK_bbox$E)+50, 
  ymn = min(ASTK_bbox$N)-50,
  ymx = max(ASTK_bbox$N)+50
  )

# Populating the raster
ASTK_raster <- rasterize(dplyr::select(ASTK_bbox, E, N), empty_raster, ASTK_bbox$AS18_72) 

# Adding arealstatistik values to deer data
deer_cleaned <- data.frame(deer_cleaned)
deer_cleaned <- deer_cleaned %>% mutate(astk = as.vector(extract(ASTK_raster, st_as_sf(deer_cleaned, coords = c("x", "y")) , 'simple')))

deer_cleaned$astk[is.na(deer_cleaned$astk)] <- 0
```

```{r}
raster_vis <- as.data.frame(ASTK_raster, xy = TRUE)

ggplot()+
  # geom_sf(data = canton_zh, fill = alpha("white", .1))+
  geom_sf(data = bbox_sf)+
  geom_raster(data = raster_vis, aes(x = x, y = y, alpha = layer))+
  geom_path(data = deer_cleaned, aes(x = x, y = y, col = TierID), alpha = 0.3)+
  geom_point(data = deer_cleaned, aes(x = x, y = y, col = TierID), size = .1)+
  coord_sf(datum = 2056)+
  theme_minimal()+
  ggtitle("Tracking points for roe deer")
  
ggplot()+
  geom_raster(data = raster_vis, aes(x = x, y = y, alpha = layer))+
  coord_sf(datum = 2056)+
  geom_point(data = deer_cleaned, aes(x,y, color = as.factor(astk)))+
  geom_path(data = deer_cleaned, aes(x = x, y = y), alpha = 0.3)+
  theme_minimal()+
  ggtitle("Tracking points for roe deer with the Arealstatistik")
```

### Methods
```{r}
# Filte roe deer 1,2 and 3 and change coordinates to sf object
deer_01 <- deer_cleaned %>% filter(TierID == "RE01") %>% na.omit()
deer_02 <- deer_cleaned %>% filter(TierID == "RE02") %>% na.omit()
deer_03 <- deer_cleaned %>% filter(TierID == "RE03") %>% na.omit()

deer1_3 <- rbind(deer_01, deer_02, deer_03)

deer_01_sf <- deer_01 %>% st_as_sf(coords = c("x", "y"), crs = 2056, remove = FALSE)
deer_02_sf <- deer_01 %>% st_as_sf(coords = c("x", "y"), crs = 2056, remove = FALSE)
deer_03_sf <- deer_01 %>% st_as_sf(coords = c("x", "y"), crs = 2056, remove = FALSE)

```

```{r}
#Overview over changepoints in speed for the roe deer
CP_value01 = cpt.mean(deer_01$speed, method= "BinSeg") # Using method "binseg" cause it is quick but only approximate
CP_value01
cpts(CP_value01)

plot(CP_value01,xlab="Date", ylab = "Speed [m/s]" ,main =" Change points in the speed of roe deer 1")

CP_value02 = cpt.mean(deer_02$speed, method= "BinSeg") # Using method "binseg" cause it is quick but only approximate
CP_value02
cpts(CP_value02)

plot(CP_value02,xlab="Date", ylab = "Speed [m/s]" ,main =" Change points in the speed of roe deer 2")

CP_value03 = cpt.mean(deer_03$speed, method= "BinSeg") # Using method "binseg" cause it is quick but only approximate
CP_value03
cpts(CP_value03)

plot(CP_value03,xlab="Date", ylab = "Speed [m/s]" ,main =" Change points in the speed of roe deer 3")
```
Surprisingly we do not set any changepoints in the speed values for the roe deer.

```{r}
### Forecasting speed 
speed_forcasting <- function(file){
  deer_mean_speed = file %>% 
  group_by(DatetimeUTC) %>% 
  summarise(speed = mean(speed, na.rm = TRUE)) %>% 
  ungroup() %>% 
  rename(ds = DatetimeUTC, y = speed)

m.s= prophet(df = deer_mean_speed)

future.s = make_future_dataframe(m = m.s, 
                               periods = 15, 
                               freq = "day")

future.speed = stats::predict(m.s, future.s)

print(plot(m.s,future.speed) +
  theme_bw()+
  theme(axis.text = element_text(size = 12), axis.title = element_text(size = 14))+
  labs(x = NULL, y = expression(Roe~deer~speed~(m/s))))

dyplot.prophet(x = m.s, fcst = future.speed,uncertainty = TRUE )

prophet_plot_components(m = m.s, fcst = future.speed)
}

deer_01_s.f <- speed_forcasting(deer_01)
deer_02_s.f <- speed_forcasting(deer_02)
deer_03_s.f <- speed_forcasting(deer_03)
```
### Home range estimation
```{r}
min_max <- function(file){ 
  min_speed = tail(file[[1]][["data"]][["trend_lower"]], n = 1) 
  max_speed = tail(file[[1]][["data"]][["trend_upper"]], n = 1)
  min_distance = (min_speed * 86400)
  max_distance = (max_speed * 86400)
  df <- data.frame (min_HR = min_speed,
                    max_HR = max_speed,
                    min_distance = min_distance,
                    max_distance = max_distance)
}

deer01_HR <- min_max(deer_01_s.f)
deer02_HR <- min_max(deer_02_s.f)
deer03_HR <- min_max(deer_03_s.f)

#Home range distance definition
homerange <- function(file, f1){
  last_point_x = last(file$x)
  last_point_y = last(file$y)
  
  buffer_point <- data.frame(x = last_point_x, y = last_point_y)
  buffer_point <- buffer_point %>% st_as_sf(coords = c("x", "y"), crs = 2056, remove = FALSE)
  
  min_buffer <- buffer_point %>% st_buffer(dist= f1$min_distance)
  max_buffer <- buffer_point %>% st_buffer(dist= f1$max_distance)
}

homerange_deer1 <- homerange(deer_01, deer01_HR)
homerange_deer2 <- homerange(deer_02, deer02_HR)
homerange_deer3 <- homerange(deer_03, deer03_HR)

ggplot()+
  geom_path(data = deer1_3, aes(x,y, col = TierID))+
  geom_point(data = deer1_3, aes(x,y, col = TierID))+
  geom_sf(data = homerange_deer1, col = "red", alpha = .1)+
  geom_sf(data = homerange_deer2, col = "darkgreen", alpha = .1)+
  geom_sf(data = homerange_deer3, col = "blue", alpha = .1)+
  ggtitle("Predicted homerange for Roe deer 1,2 and 3 in the next 24 hours")+
  xlab("X")+
  ylab("Y")+
  theme_minimal()+
  coord_sf(datum = 2056)
```

### Linear predictive model
```{r LM}
# https://cran.r-project.org/web/packages/jtools/vignettes/effect_plot.html
#Linear approach
fit <- lm(speed ~ sex + timelag + x + y + steplength + astk + TierID, data = deer_cleaned)

#predicts the future values
s.astk.model <- lm(speed~astk,data = deer_cleaned)
variable_astk <-data.frame(astk=c(deer_cleaned$astk))
p1 <- predict(s.astk.model, newdata = variable_astk)
p2 <- predict(s.astk.model, newdata = variable_astk,interval = 'confidence')


effect_plot(s.astk.model, pred = astk, interval = TRUE, plot.points = TRUE)

```

## Results
```{r}
summ(fit)
effect_plot(fit, pred = astk, interval = TRUE, plot.points = TRUE)

fit_poly <- lm(speed ~ sex + timelag + x + y + steplength + poly(astk, 2) + TierID, data = deer_cleaned)
effect_plot(fit_poly, pred = astk, interval = TRUE, plot.points = TRUE)


effect_plot(fit, pred = astk, interval = TRUE, plot.points = TRUE,
            jitter = .2)

summ(s.astk.model)

plot(p1)
plot(p2)
```
## Discussion


## Bibliography
